### 2013-05-19 draft.tex
fixed some spelling mistakes, added reference to veizer figure

Despite the importance of paleoclimate and more than half a century of investigations on the topic, the temperature history of Earth's oceans remains poorly constrained. The most widely-used proxy for past ocean temperatures is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate \deltao O thermometer, which was developed by Harold Urey in the 1940s \citep{Urey1947}. This carbonate \deltao O proxy has two significant sources of complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary envrionmental signature. Second, the measured \deltao O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the \deltao O of the seawater. The \deltao O of seawater can be affected by several factors, including evaporation, mixing with freshwater, and changes in the size of continental glaciers, all of which are difficult to quantify in the geologic record. The 
\deltao O of the ocean has also been hypothesized to change gradually but substantially through Earth's history \citep{Jaffres2007,Veizer1999}.

Gradual changes in the \deltao O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the \deltao O of seawater. The long-term average \deltao O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the \deltao O of ocean water by equilibrating the water with the average mantle \deltao O of 5.7$\permil$ \citep{Gregory1991}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the \deltao O of seawater by preferentially moving $^{18}$O from water into rocks. The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995,Tajika1993,Wallmann2001} and the magnitude of contintental weathering, which would lead to slow shifts in the \deltao O of the ocean. The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controvery in geochemistry.

@@

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected. In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease. In constrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.

@@

Electron backscatter diffraction (EBSD) provides an independent measure of recrystallization by mapping the orientation of calcite crystals in a thin section. This allows more detailed characterization of calcite fabrics than is possible using optical microscopy, and in the case of brachiopods, comparison to pristine modern examples \citep{PerezHuerta2007}. Two thin sections, one rugose coral and one brachiopod, were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under variable pressure vacuum, with an electron high tension of 20 kV. The stage was tilted 70$^{\circ}$. The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame and a step size of 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod. The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite. The diffracted beam then interacts with the EBSD detector, depdending on the orientation of the calcite crystal. The data were analyzed using software from Oxford instruments.

Flakes of the secondary fibrous layer from over ten different brachiopods were also imaged to further inspect the preservation of original fibrous calcite texture. The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick. The coated flakes were imaged using using the secondary electron detector in the same SEM as above, under high vacuum. The working distance was $\sim 6$ mm and the electron high tension was 10 kV.

@@

The use of wavelength-dispersive spectroscopy in combination with optical microscopy and EBSD allows characterization of the trace metal contents of specific calcite textures. By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set. The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer (E-probe). Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite. Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel. Quantitative spot measurements were made using the JEOL quantitiative analysis program.

@@

$\Delta_{47-\text{RF-AC}}$ values were converted to temperature using the \cite{Ghosh2006} calibration line, which is projected into the absolute reference frame in equation 9 in \cite{Dennis2011}. The \deltao O of water was calcuated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.

@@

% array. probably not solid-state reordering, so we interpret rock-buffered diagenesis
% comparisons support hypothesis that low temp most pristine, EBSD supports, metals dont negate
% most pristine samples population at 34, one at 30, resample not replicate, no big deal!
% only upper estimates, but probably pretty close
% Silurian warm and similar in d18O to today. matches previous clumped studies
% resolves veizer controversy
% implications of warm oceans for Ireviken climate rule out big changes, can't resolve small
% implications of constant d18O?? relative amounts constant. either both constant or coupled.

@@

A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates. We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples. We are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine samples. However, micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, comprising a maximum of 5\% of the material. Assuming linear mixing, which is not technically correct but close enough for our purposes, if 5\% of the sample material was recrystallized at 100\degrees C and the rest of the sample material retained an original temperature of 30\degrees C, the measured temperature of the sample would be $\sim34$\degrees C. Therefore, the constraints provided by the EBSD mapping of the bachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.

@@

Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates. Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($>50$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

@@

\vspace{0.7cm} \noindent \large \textbf{Acknowldgements} \normalsize

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.6\textwidth]{Fig_veizer-01.jpg}
\caption{In gray, the \deltao O$_{\text{PDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}. In black are two different hypotheses for the trend in the \deltao O$_{\text{SMOW}}$ of the oceans through time.}
\label{veizer}
\end{figure}

>>>>

Despite the importance of paleoclimate and more than half a century of investigations on the topic, the temperature history of Earth's oceans remains poorly constrained. The most widely-used proxy for past ocean temperatures is the carbonate-water oxygen isotope exchange thermometer, also known as the carbonate \deltao O thermometer, which was developed by Harold Urey in the 1940s \citep{Urey1947}. This carbonate \deltao O proxy has two significant sources of complications. First, as with many paleoenvironmental proxies, it is often difficult to distinguish the effects of diagenetic alteration from the primary environmental signature. Second, the measured \deltao O of carbonates depends not only on the temperature of the seawater from which the carbonates precipitated, but also on the \deltao O of the seawater. The \deltao O of seawater can be affected by several factors, including evaporation, mixing with freshwater, and changes in the size of continental glaciers, all of which are difficult to quantify in the geologic record. The 
\deltao O of the ocean has also been hypothesized to change gradually but substantially through Earth's history \citep{Jaffres2007,Veizer1999}.

Gradual changes in the \deltao O of seawater on 100-million-year timescales seem plausible given the current knowledge of the controls on the \deltao O of seawater. The long-term average \deltao O of all the water on Earth's surface is almost entirely controlled by the balance between two exchange processes: exchange of oxygen between water and rocks during high-temperature alteration of the seafloor near mid-ocean ridges, and exchange of oxygen between water and rocks during low-temperature weathering of oceanic and continental crust. High-temperature exchange processes have small oxygen isotope fractionation factors and tend to increase the \deltao O of ocean water by equilibrating the water with the average mantle \deltao O of 5.7$\permil$ \citep{Gregory1991}. Low-temperature exchange processes have larger oxygen isotope fractionation factors and tend to decrease the \deltao O of seawater by preferentially moving $^{18}$O from water into rocks. The relative magnitudes of high- and low-temperature weathering are likely to have changed over the Phanerozoic due to uncoupled changes in the seafloor spreading rate \citep{Fornari1995,Tajika1993,Wallmann2001} and the magnitude of contintental weathering, which would lead to slow shifts in the \deltao O of the ocean. The best geologic evidence for gradual change in seawater \deltao O, however, is the subject of a long-standing controversy in geochemistry.

@@

In addition to being highly sensitive to diagenetic alteration, carbonate clumped isotopes are also diagnostic of diagenetic regimes, ultimately allowing identification of the most pristine endmember of a sample set. Clumped isotopes can distinguish between diagenesis in the presence of low and high water-to-rock ratios (rock-buffered and water-buffered diagenesis, respectively) and diagenesis in the presence of meteoric water. As carbonates are buried deeper and the temperature increases, the equilibrium oxygen isotope fractionation between water and carbonate grows smaller, preferentially driving $^{18}$O into the water. During rock-buffered diagenesis, this causes the \deltao O of the water to increase, while the \deltao O of the altered carbonate is not affected. In a water-buffered diagenetic regime, the \deltao O of the water does not change, so increasing temperature instead causes the \deltao O of the altered carbonate to decrease. In contrast, diagenesis in the presence of meteoric water usually occurs when the sample is near the Earth's surface, so the clumped isotope temperature of the altered carbonate is cool, but the \deltao O of the carbonate and the water is diagnostically low. The ability of clumped isotopes to constrain the conditions to which fossils have been exposed after burial allows identification of the best-preserved samples from within a set of samples that traditionally would all be considered pristine \citep{Eiler2011}. We demonstrate here that this distinction has important implications for the estimation of the temperature and \deltao O of oceans in the past.

@@

Electron backscatter diffraction (EBSD) provides an independent measure of recrystallization by mapping the orientation of calcite crystals in a thin section. This allows more detailed characterization of calcite fabrics than is possible using optical microscopy, and in the case of brachiopods, comparison to pristine modern examples \citep{PerezHuerta2007}. Two thin sections, one rugose coral and one brachiopod, were analyzed using a Zeiss 1550VP Field Emission Scanning Electron Microscrope (SEM) under variable pressure vacuum, with an electron high tension of 20 kV. The stage was tilted 70$^{\circ}$. The AZtec program was used in $4\times4$ binning mode, with frame averaging over one frame and a step size of 2 $\mu$m for the rugose coral and 1.2 $\mu$m for the brachiopod. The electron beam interacts with the crystal and is diffracted in a characteristic pattern for calcite. The diffracted beam then interacts with the EBSD detector, depending on the orientation of the calcite crystal. The data were analyzed using software from Oxford instruments.

Flakes of the secondary fibrous layer from over ten different brachiopods were also imaged to further inspect the preservation of original fibrous calcite texture. The rinsed and dried flakes were prepared for imaging by coating them with a layer of carbon 12 nm thick. The coated flakes were imaged using the secondary electron detector in the same SEM as above, under high vacuum. The working distance was $\sim 6$ mm and the electron high tension was 10 kV.

@@

The use of wavelength-dispersive spectroscopy in combination with optical microscopy and EBSD allows characterization of the trace metal contents of specific calcite textures. By comparing the trace metal concentrations of primary textures to those of diagenetic textures, we can identify the trace metal signature of diagenetic alteration in our sample set. The same thin sections as above were coated with a layer of carbon 15 nm thick, and then analyzed with wavelength-dispersive spectroscopy using the JEOL JXA-8200 Electron Probe Micro-analyzer (E-probe). Using the JEOL map analysis program, the beam was calibrated on crystals of dolomite, siderite, strontianite and rhodochrosite. Qualitative maps of Ca, Mg, Fe, Sr and Mn concentration were made using a current of 200 nAmps, voltage of 15 kV, and a counting time of 80 milliseconds per pixel. Quantitative spot measurements were made using the JEOL quantitative analysis program.

@@

$\Delta_{47-\text{RF-AC}}$ values were converted to temperature using the \cite{Ghosh2006} calibration line, which is projected into the absolute reference frame in equation 9 in \cite{Dennis2011}. The \deltao O of water was calculated from the temperature and the measured \deltao O of the sample using the temperature-dependent fractionation, $\alpha$, of oxygen isotopes between water and calcite, $1000\ln\alpha=18.03\left(10^3\text{T}^{-1}\right)-32.42$, where T is temperature in degrees Kelvin \citep{Kim1997}.

@@

@@

A second caveat of our estimates of the temperature and \deltao O of Silurian tropical oceans is that they are technically upper estimates. We are confident that our sample suite was subjected to a diagenetic regime that increased the temperature and \deltao O of water recorded by the samples. We are not able to exclude the possibility that diagenesis affected all the samples in our study, even the most pristine samples. However, micro-scale studies of the calcite fabric of the lowest-temperature brachiopods indicate that late-phase calcite is only a trace constituent of these samples, comprising a maximum of 5\% of the material. Assuming linear mixing, which is not technically correct but close enough for our purposes, if 5\% of the sample material was recrystallized at 100\degrees C and the rest of the sample material retained an original temperature of 30\degrees C, the measured temperature of the sample would be $\sim34$\degrees C. Therefore, the constraints provided by the EBSD mapping of the brachiopod shell calcite demonstrate that our estimate of the temperature of the tropical Silurian ocean is likely reliable to within 4\degrees C.

@@

Our results, along with those of previous clumped isotope studies \citep{Came2007,Dennis2013,Finnegan2011}, provide consistent explanations for the low \deltao O values of early Paleozoic carbonates (Figure \ref{Veizer}). Carbonate clumped isotopes constrain the contribution of a change in seawater \deltao O to only a few permil of the observed trend, ruling out a value of $-8\permil$ for the \deltao O$_{\text{SMOW}}$ of early Paleozoic oceans \citep{Grossman2012}. Our results also serve as a reminder that many of the fossils preserved in the geologic record lived in tropical oceans. The \deltao O of carbonate in equilibrium with tropical ocean water should not be compared to the \deltao O of carbonate in equilibrium with water of average modern sea surface temperature, because tropical oceans are much warmer than the average sea surface temperature. Clumped isotopes additionally demonstrate that extremely hot ($>50$\degrees C) ocean temperatures cannot be the cause of the depleted \deltao O values of early Paleozoic carbonates. Finally, clumped isotopes, as a highly sensitive indicator of diagenetic alteration, demonstrate that even among samples with original calcite fabric and pristine trace metal signatures, there is a large range in preservation quality. This means that the average fossil calcite in almost any sample suite is likely altered from its primary composition, and that diagenetic alteration is a key contributor to the low \deltao O values of early Paleozoic carbonates.

@@

\vspace{0.7cm} \noindent \large \textbf{Acknowledgements} \normalsize

@@

\begin{figure}[htb]
\centering
\includegraphics[width=0.6\textwidth]{Fig_veizer-01.jpg}
\caption{In gray, the \deltao O$_{\text{PDB}}$ of apparently well-preserved skeletal calcites, compiled by \cite{Veizer1999}. In black are two different hypotheses for the trend in the \deltao O$_{\text{SMOW}}$ of the oceans through time.}
\label{Veizer}
\end{figure}
